\pkg{ggenealogy}: An \proglang{R} Package for Visualizing Genealogical Data

### 2016-11-12 article.tex
plotPath(), plotPathOnAll() input changed
all but animate in replication script

\begin{figure}[H]
    \centering
    \fbox{\includegraphics[width=0.65\textwidth]{kinshipFig}}
    \caption{Example pedigree chart from the \pkg{kinship2} package, where the vertical axis denotes generation count. We superimposed green-highlighted individual 215 for explanatory purposes. As an offspring of a parent-child relationship, individual 215 is both a second and third generation individual. Hence, it should be displayed twice on the vertical axis, once for each of its generation counts. However, most standard pedigree tools only allow for an individual to be displayed once. In the \pkg{kinship2} package, individuals can indeed be displayed more than once. However, each child must have zero or two parents (one male and one female). These restrictions make it impossible to plot genealogical data by generation count in cases where there are many inter-generational breedings.
    \label{fig:kinshipFig}
\end{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> install.packages("ggenealogy")
R> library("ggenealogy")
R> data("sbGeneal")
R> str(sbGeneal)
\end{CodeInput}
\begin{CodeOutput}
'data.frame':	390 obs. of  5 variables:
 $ child       : chr  "5601T" "Adams" "A.K." "A.K. (Harrow)" ...
 $ date        : num  1981 1948 1910 1912 1968 ...
 $ yield       : int  NA 2734 NA 2665 NA 2981 2887 2817 NA NA ...
 $ date.imputed: logi  TRUE FALSE TRUE FALSE FALSE FALSE ...
 $ parent      : chr  "Hutcheson" "Dunfield" NA "A.K." ...
\end{CodeOutput}
\end{CodeChunk}
\subsection{Academic genealogy of statisticians}
The \pkg{ggenealogy} package also comes with an academic genealogy of statisticians; this dataset is in the form of a data frame with 8165 rows and six columns. To develop this later dataset, we contacted the \citealt{mgp}, a web-based database for the genealogy of academic mathematicians. This database, which currently contains almost 200,000 entries, is a service of the North Dakota State University Department of Mathematics and the American Mathematical Society. The Mathematics Genealogy Project contact provided us a Structured Query Language (SQL) export, and we used PostgreSQL to query the database (\citealt{psql}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("statGeneal")
R> dim(statGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] 8165    6
\end{CodeOutput}
\begin{CodeInput}
R> colnames(statGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] "child"   "parent"  "date"    "country" "school"  "thesis" 
\end{CodeOutput}
\end{CodeChunk}

@@

However, some \pkg{ggenealogy} plots also make use of quantitative variable values associated with individuals in the genealogy. For these plots, the input data frame should also contain a third quantitative variable column. Each case in this quantitative variable column can be of type numeric, type NA, or type ``". In both example data files, this column is labeled ``date," and each case in that column can be of type numeric, type NA, or type ``". At this point, any \pkg{ggenealogy} plot can be used.

@@

With soybean lineages, it may be useful for soybean breeders to track how two varieties are related to each other via parent-child relationships. Then, any dramatic changes in yield and other measures of interest between the two varieties can be traced across their genetic timeline. The \pkg{ggenealogy} package allows users to select two varieties of interest, and determine the shortest pathway of parent-child relationships between them, using the \code{getPath()} function. This will return a list that contains the variety names and their years in the path.

\begin{CodeChunk}
\begin{CodeInput}
R> pathTN <- getPath("Tokyo", "Narow", sbIG, sbGeneal)
R> pathTN
\end{CodeInput}
\begin{CodeOutput}
$pathVertices
[1] "Tokyo"    "Volstate" "Jackson"  "R66-873"  "Narow"   
$dateVertices
[1] "1907"   "1942"   "1954.5" "1971.5" "1985"  
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{Code}
R> plotPath(pathTN)
\end{Code}

@@

Next, we can run the same set of functions on a different pair of varieties. First, a call to the \pkg{ggenealogy} function \code{getDate()} indicates that variety \code{Bedford} was released in 1978 and variety \code{Zane} in 1985.

\begin{CodeChunk}
\begin{CodeInput}
R> getDate("Bedford", sbGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] 1978
\end{CodeOutput}
\begin{CodeInput}
R> getDate("Zane", sbGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] 1985
\end{CodeOutput}
\end{CodeChunk}

We can then create a plot showing the shortest path between these two varieties of interest. As this is a longer path, we may also consider setting the \code{fontFace} variable of the \code{plotPath()} to a value of 2, indicating we wish to boldface the two varieties of interest.

\begin{Code}
R> pathBZ <- getPath("Bedford", "Zane", sbIG, sbGeneal)
R> plotPath(pathBZ, fontFace = 2)
\end{Code}

@@

\begin{Code}
R> plotPathOnAll(pathTN, sbGeneal, sbIG, bin = 3, pathEdgeCol = "red",
+    nodeSize = 2.5, pathNodeSize = 4) + ggplot2::theme(axis.text = ggplot2::
+    element_text(size = 12), axis.title = ggplot2::element_text(size = 12))
\end{Code}

@@

\begin{Code}
R> plotPathOnAll(pathTN, sbGeneal, sbIG, bin = 6, pathEdgeCol =
+    "seagreen2", nodeSize = 1, pathNodeSize = 3) + ggplot2::xlab("Year")
\end{Code}

@@

Notice from Figure~\ref{fig:plotTNBin3} that the default horizontal axis label for the \texttt{plotPath()} method has a value of ``Date". Given that the ``date" variable in this example dataset is on the timescale of years, we wanted to change the default value of the horizontal axis label to ``Year". We did this in the code above for Figure~\ref{fig:plotTNBin6} by appending the \texttt{ggplot2::xlab()} function.

@@

\begin{Code}
>R varieties <- c("Brim", "Bedford", "Calland", "Dillon", "Hood", "Narow",
+    "Pella", "Tokyo", "Young", "Zane")
>R plotDegMatrix(varieties, sbIG, sbGeneal, "Variety", "Variety", "Degree") +
+    ggplot2::scale_fill_continuous(low = "white", high = "darkgreen") +
+    ggplot2::theme(legend.title = ggplot2::element_text(size = 15),
+    legend.text = ggplot2::element_text(size = 15))
\end{Code}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> statIG <- dfToIG(statGeneal)
R> pathCB <- getPath("David Cox", "Petra Buzkova", statIG, statGeneal,
+    isDirected = FALSE)
R> plotPath(pathCB, fontFace = 4) + ggplot2::xlab("Year") +
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 10), axis.title =
+    ggplot2::element_text(size = 10)) + + ggplot2::scale_x_continuous(expand
+    = c(.1, .2))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plotPathOnAll(pathCB, statGeneal, statIG, bin = 200)
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 8), axis.title =
+    ggplot2::element_text(size = 8)) + ggplot2::scale_x_continuous
+    (expand = c(.1, .2)) + ggplot2::xlab("Year")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plotPathOnAll(pathCB, statGeneal, statIG, bin = 200, nodeSize = .5,
+    pathNodeSize = 2.5, nodeCol = "darkgray", edgeCol = "lightgray") +
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 8), axis.title =
+    ggplot2::element_text(size = 8)) + ggplot2::scale_x_continuous
+    (expand = c(.1, .2)) + ggplot2::xlab("Year")
\end{CodeInput}
\end{CodeChunk}

>>>>

\begin{figure}[H]
    \centering
    \fbox{\includegraphics[width=0.65\textwidth]{kinshipFig}}
    \caption{Example pedigree chart from the \pkg{kinship2} package, where the vertical axis denotes generation count. We superimposed green-highlighted individual 215 for explanatory purposes. As an offspring of a parent-child relationship, individual 215 is both a second and third generation individual. Hence, it should be displayed twice on the vertical axis, once for each of its generation counts. However, most standard pedigree tools only allow for an individual to be displayed once. In the \pkg{kinship2} package, individuals can indeed be displayed more than once. However, each child must have zero or two parents (one male and one female). These restrictions make it impossible to plot genealogical data by generation count in cases where there are many inter-generational breedings.}
    \label{fig:kinshipFig}
\end{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> install.packages("ggenealogy")
R> library("ggenealogy")
R> data("sbGeneal")
R> str(sbGeneal)
\end{CodeInput}
\begin{CodeOutput}
'data.frame':	390 obs. of  5 variables:
 $ child      : chr  "5601T" "Adams" "A.K." "A.K. (Harrow)" ...
 $ devYear    : num  1981 1948 1910 1912 1968 ...
 $ yield      : int  NA 2734 NA 2665 NA 2981 2887 2817 NA NA ...
 $ yearImputed: logi  TRUE FALSE TRUE FALSE FALSE FALSE ...
 $ parent     : chr  "Hutcheson" "Dunfield" NA "A.K." ...
\end{CodeOutput}
\end{CodeChunk}

\subsection{Academic genealogy of statisticians}
The \pkg{ggenealogy} package also comes with an academic genealogy of statisticians; this dataset is in the form of a data frame with 8165 rows and six columns. To develop this later dataset, we contacted the \citealt{mgp}, a web-based database for the genealogy of academic mathematicians. This database, which currently contains almost 200,000 entries, is a service of the North Dakota State University Department of Mathematics and the American Mathematical Society. The Mathematics Genealogy Project contact provided us a Structured Query Language (SQL) export, and we used PostgreSQL to query the database (\citealt{psql}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("statGeneal")
R> dim(statGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] 8165    6
\end{CodeOutput}
\begin{CodeInput}
R> colnames(statGeneal)
\end{CodeInput}
\begin{CodeOutput}
[1] "child"    "parent"   "gradYear" "country"  "school"   "thesis" 
\end{CodeOutput}
\end{CodeChunk}

@@

However, some \pkg{ggenealogy} plots also make use of quantitative variable values associated with individuals in the genealogy. For these plots, the input data frame should also contain a third quantitative variable column. Each case in this quantitative variable column can be of type numeric, type NA, or type ``". In the first example dataset, columns that could be used for this purpose include ``devYear", ``yield", and ``yearImputed"; in the second example dataset, the column that could be used for this purpose is ``gradYear". At this point, any \pkg{ggenealogy} plot can be used.

@@

With soybean lineages, it may be useful for soybean breeders to track how two varieties are related to each other via parent-child relationships. Then, any dramatic changes in yield and other measures of interest between the two varieties can be traced across their genetic timeline. The \pkg{ggenealogy} package allows users to select two varieties of interest, and determine the shortest pathway of parent-child relationships between them, using the \code{getPath()} function. This will return a list that contains the path, along with the variety name and quantitative variable value of interest for each variety in the path. For this example, we will use the development year (from the column ``devYear") as our quantitative variable.

\begin{CodeChunk}
\begin{CodeInput}
R> pathTN <- getPath("Tokyo", "Narow", sbIG, sbGeneal, "devYear")
R> pathTN
\end{CodeInput}
\begin{CodeOutput}
$pathVertices
[1] "Tokyo"    "Volstate" "Jackson"  "R66-873"  "Narow"   
$variableVertices
[1] "1907"   "1942"   "1954.5" "1971.5" "1985"   
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{Code}
plotPath(pathTN, "devYear")
\end{Code}

@@

Next, we can run the same set of functions on a different pair of varieties. First, we can call the \pkg{ggenealogy} function \code{getVariable()} using the input quantitative variable of development year. This indicates that variety \code{Bedford} was released in 1978 and variety \code{Zane} in 1985.

\begin{CodeChunk}
\begin{CodeInput}
R> getVariable("Bedford", sbGeneal, "devYear")
\end{CodeInput}
\begin{CodeOutput}
[1] 1978
\end{CodeOutput}
\begin{CodeInput}
R> getVariable("Zane", sbGeneal, "devYear")
\end{CodeInput}
\begin{CodeOutput}
[1] 1985
\end{CodeOutput}
\end{CodeChunk}

We can then create a plot showing the shortest path between these two varieties of interest. As this is a longer path, we may also consider setting the \code{fontFace} variable of the \code{plotPath()} to a value of 2, indicating we wish to boldface the two varieties of interest. In addition, as is the case with plotting tools in \pkg{ggenealogy}, we can append \pkg{ggplot2} syntax. In this case, we may wish to hard code the x-axis label from its default value of ``devYear" (the inputted quantitative variable column name) to the more readable ``Development Year".

\begin{Code}
R> pathBZ <- getPath("Bedford", "Zane", sbIG, sbGeneal, "devYear")
R> plotPath(pathBZ, "devYear", fontFace = 2) + ggplot2::xlab("Development Year")
\end{Code}

@@

\begin{Code}
R> plotPathOnAll(pathTN, sbGeneal, sbIG, "devYear", bin = 3, pathEdgeCol = "red",
+    nodeSize = 2.5, pathNodeSize = 4) + ggplot2::theme(axis.text = ggplot2::
+    element_text(size = 12), axis.title = ggplot2::element_text(size = 12))
\end{Code}

@@

\begin{Code}
R> plotPathOnAll(pathTN, sbGeneal, sbIG, "devYear", bin = 6, pathEdgeCol =
+    "seagreen2", nodeSize = 1, pathNodeSize = 3) + ggplot2::xlab("Development Year")
\end{Code}

@@

Notice again from Figure~\ref{fig:plotTNBin3} that the default horizontal axis label for the \texttt{plotPath()} method has a value of ``devYear". We wanted to change the default value of the horizontal axis label to ``Development Year". We did this in the code above for Figure~\ref{fig:plotTNBin6} by appending the \texttt{ggplot2::xlab()} function.

@@

\begin{Code}
>R varieties <- c("Brim", "Bedford", "Calland", "Dillon", "Hood", "Narow",
+    "Pella", "Tokyo", "Young", "Zane")
>R plotDegMatrix(varieties, sbIG, sbGeneal, "devYear") +
+    ggplot2::scale_fill_continuous(low = "white", high = "darkgreen") +
+    ggplot2::theme(legend.title = ggplot2::element_text(size = 15),
+    legend.text = ggplot2::element_text(size = 15)) +
+    ggplot2::labs(x = "Variety", y = "Variety")
\end{Code}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> statIG <- dfToIG(statGeneal)
R> pathCB <- getPath("David Cox", "Petra Buzkova", statIG, statGeneal,
+    "gradYear", isDirected = FALSE)
R> plotPath(pathCB, "gradYear", fontFace = 4) + ggplot2::xlab("Graduation Year") +
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 10), axis.title =
+    ggplot2::element_text(size = 10)) + + ggplot2::scale_x_continuous(expand
+    = c(.1, .2))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plotPathOnAll(pathCB, statGeneal, statIG, "gradYear", bin = 200)
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 8), axis.title =
+    ggplot2::element_text(size = 8)) + ggplot2::scale_x_continuous
+    (expand = c(.1, .2)) + ggplot2::xlab("Graduation Year")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plotPathOnAll(pathCB, statGeneal, statIG, "gradYear", bin = 200, nodeSize = .5,
+    pathNodeSize = 2.5, nodeCol = "darkgray", edgeCol = "lightgray") +
+    ggplot2::theme(axis.text = ggplot2::element_text(size = 8), axis.title =
+    ggplot2::element_text(size = 8)) + ggplot2::scale_x_continuous
+    (expand = c(.1, .2)) + ggplot2::xlab("Graduation Year")
\end{CodeInput}
\end{CodeChunk}
